# Data Format

The `lmd` data format is described on page 4-5 of the
[https://www.beckmancoulter.com/wsrportal/ajax/downloadDocument/773231AD.pdf?autonomyId=TP_DOC_32383&documentName=773231AD.pdf&documentAccessType=PUBLIC](Gallios
Manual). [../refs/gallios_lmd_format.pdf](Excerpt). From that source, the `lmd`
format is described as:

> Flow Cytometric Standard (FCS) file format for listmode and histogram
> files. Listmode files contain an FCS 2.0 dataset followed by an FCS 3.0
> dataset. 

The structure is:

|----------------------|
| FCS 2.0              |
|----------------------|
| Acquisition Protocol |
|----------------------|
| FCS 3.0              |
|----------------------|

# Getting Started
## Setup
Initial setup requires installing the necessary packages from Bioconductor.
This step is necessary only when initially installing or upgrading the
packages, you don't normally need to run it:

```{r, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite()
biocLite("BiocUpgrade")
biocLite("flowCore")                    # core functions
biocLite("flowViz")                     # for plotting
biocLite("flowFit")                     # model fitting for histograms
```

After the packages are installed, barring upgrades, this is all that you
need: 
```{r}
library(flowCore)
```
## Importing Data
### flowFrame

Load the data with `read.FCS`, producing individual `flowFrame` objects.
The argument `dataset` indicates which format to extract: `1` selects FCS
Version 2.0, `2` selections version 3.0.

```{r}
lmd1 <- read.FCS("test-data/LMD good/188-15_+_rad_2015-04-20_647.LMD",
                 dataset = 1, alter.names = TRUE)
lmd2 <- read.FCS("test-data/LMD good/188-15_+_rad_2015-04-20_647.LMD",
                 dataset = 2) 
lmd1
summary(lmd1)
```

The result is an object of class `flowFrame`:

>      This class represents the data contained in a FCS file or similar
>      data structure. There are three parts of the data:
>
>       1. a numeric matrix of the raw measurement values with
>          ‘rows=events’ and ‘columns=parameters’
>
>       2. annotation for the parameters (e.g., the measurement
>          channels, stains, dynamic range)
>
>       3. additional annotation provided through keywords in the FCS
>          file

There are three slots:

1. `exprs`: a matrix with events/cells as rows, channels as columns
2. `parameters`": an `AnnotatedDataFrame`, containing summaries of the
column data (access via pData(parameters(lmd2)))
3. `description`: metadata from the source file, a list of names keywords

Extract and manipulate the values from each slot with `exprs()`,
`parameters()` and `description`, respectively. Individual keywords are
accessible via `keyword`. Parameter names are accessible via `colnames` or
`featureNames` (synonyms).

### flowSet

`flowSet` is a class that contains a list of `flowFrame` objects, and some
metaData. Easiest to create directly from a folder of files:

```{r}
fs <- read.flowSet(path = "test-data/LMD good/", alter.names = TRUE,
                   dataset = 1)

```

See the help files for additional details, which I think are not crucial to
our current mission.

## Visualization

Basic plotting is provided via `flowViz`:

```{r}
library(flowViz)
par(mfrow = c(1,1))
plot(lmd1, main = "Pair-plots of Flow Channels")
plot(lmd1, c("FL3.INT.LIN", "SS.TOF.LIN"),
     main = "Scatterplot of FL3 area and SS TOF")
plot(lmd1, "FL3.INT.LIN", breaks = 256, main = "Histogram of FL3 Area")
par(mfrow = c(2,1))
plot(lmd1, "FL3.INT.LIN", breaks = 256, main = "Histogram of FL3 area, FCS2 format")
plot(lmd2, "FL3-A", breaks = 256, main = "Histogram of FL3 area, FC3 format")
```
# Histogram modelling

Extract & bin the values. Note that the highest value for each column
contains all values that are above detection thresholds, producing a
clipping artifact. Clip this value before fitting the models.

```{r}
fl3.trimmed <- exprs(lmd1)[, "FL3.INT.LIN"][exprs(lmd1)[, "FL3.INT.LIN"] <
                                            max(exprs(lmd1)[, "FL3.INT.LIN"])]
## Set one more break than the number of channels we want!
fl3.1024 <- hist(fl3.trimmed, breaks = seq(from = 0, to=1024, by = 1))
fl3.256 <- hist(fl3.trimmed, breaks = seq(from = 0, to = 1024, by = 4))
```

Convert to a dataframe, as required by `nls()`. Note that `hist()` provides
bin boundaries as `$breaks`, midpoints as `$mids`, and counts per bin as
`$counts`. The counts are our intensities, and the midpoints are the x
values in terms of the channels in the original data. However, in ModFit,
they use the channel number, rather than the channel midpoint, so to match
their output (or at least put ours on the same scale), we need to do the
same. 

```{r}
## not sure we even need to explicitly identify x, it's just an index
fl3.df <- data.frame(x = 1:length(fl3.256$counts),
                     intensity = fl3.256$counts) 
plot(fl3.df$intensity, type = 'n')
polygon(x = c(fl3.df$x, 257), y = c(fl3.df$intensity, 0),
                col = "lightgray", border = NA) 

```

Build some model components. The dependent variable in each model component
is `xx`, but is not explicitly added to the formal arguments until the
components are collected together with buildModel.

```{r}
erf <- function(x) 2 * pnorm(x * sqrt(2)) - 1

gaussDipA  <- function(a1 = a1, Ma = Ma, Sa = Sa, a2 = a2){
    a1 / (sqrt(2 * pi) * Sa) * exp(-((xx - Ma)^2)/(2 * Sa^2)) +
      a2 / (sqrt(2 * pi) * Sa * 2) * exp(-((xx - Ma * 2)^2)/(2 * (Sa * 2)^2))
}
  
gaussDipB  <- function(b1 = b1, Mb = Mb, Sb = Sb, b2 = b2){
  b1 / (sqrt(2 * pi) * Sb) * exp(-((xx - Mb)^2)/(2 * Sb^2)) +
    b2 / (sqrt(2 * pi) * Sb * 2) * exp(-((xx - Mb * 2)^2)/(2 * (Sb * 2)^2))
}

singleIter <- function(SCa, intensity){
  first.channel <- which(fl3.df$intensity > 0)[2]
  # Don't extend the model below/beyond the data
  # Modfit appears to cut off the debris slightly above the lowest data,
  # which gives a better fit. Perhaps set first.channel to 2-4? Need to
  # test this and determine best fit.
  channels = first.channel:length(intensity)
  res = vector(length = length(intensity), mode = "numeric")
  for(x in first.channel:(length(intensity) - 1)){
    for(j in (x + 1):length(intensity))
      res[x] <- res[x] + j^(1/3) * intensity[j] * 2 / (pi * j * sqrt(x/j * (1 - x/j)))
  }
  res[length(res)] <- 0
  return(SCa * res)
}

singleCut <- function(SCa, intensity){
  singleIter(SCa, intensity)
}

broadenedTrapezoid <- function(BTt1, BTt2, BTx1, BTx2, BTs1, BTs2){
  ((BTt2 - BTt1) / (BTx2 - BTx1) * (xx - BTx2) + BTt2) *
    ((erf((BTx2 - xx)/sqrt(2 * BTs2)) -
      erf((BTx1 - xx)/sqrt(2 * BTs1))) / 2)
}

buildModel <- function(models) {
  ## Model components combined by buildModel must have a function body that
  ## is composed of a single line that evaluates to the return value of the
  ## function. If multi-line bodies are required, as in the iterative
  ## components like the single cut debris model, use a wrapper function to
  ## preserve the one-line body.
  modelComponents <- vector(mode = "character", length(models))
  
  model <- function(){}
  
  for (i in seq_along(models)){
    modelComponents[i] <- as.character(body(get(models[[i]])))[2]
    formals(model) <- c(formals(model), formals(models[i]))
  }
  
  modelParsed <- parse(text = paste(modelComponents, collapse = " + "))
  body(model) <- modelParsed
  
  formals(model) <- c(alist(xx = ), formals(model) )
  
  return(model)
}

oneSample <- buildModel(c("gaussDipA", "gaussDipB"))

oneSampleSC <- buildModel(c("gaussDipA", "gaussDipB", "singleCut"))

```

## Nonlinear regression

```{r}

## For eyeballing the starting values:
## tmp <- oneSampleSC(Ma = 97, a1 = 3000, Sa = 5, a2 = 500,
##                    Mb = 135, b1 = 5000, Sb = 5, b2 = 10,
##                    SCa = 0.1, intensity = fl3.df$intensity,
##                    xx = fl3.df$x)

m3 <-
  nls(intensity ~
        oneSampleSC(a1 = a1, Ma = Ma, Sa = Sa, a2 = a2,
                    b1 = b1, Mb = Mb, Sb = Sb, b2 = b2, SCa = SCa,
                    intensity = intensity, xx = x),
      data = fl3.df,
      start = list(Ma = 97, a1 = 3000, Sa = 5, a2 = 500,
                   Mb = 135, b1 = 5000, Sb = 5, b2 = 10,
                   SCa = 0.1))

## The following doesn't work yet, I suspect because the optimizer attempts
## negative values for the broadened trapezoid heights (the true values are
## very close to zero). Need to figure out how to constrain values in the
## nls procedure before I can use this.
m4 <-
  nls(intensity ~
        oneSampleSCBT(a1 = a1, Ma = Ma, Sa = Sa, a2 = a2,
                    b1 = b1, Mb = Mb, Sb = Sb, b2 = b2, SCa = SCa,
                    intensity = intensity,
                    BTt2 = BTt2, BTt1 = BTt1, BTx1 = Ma, BTx2 = 2 * Ma,
                    BTs1 = BTs1, BTs2 = BTs2,
                    xx = x), 
      data = fl3.df,
      start = list(Ma = 400, a1 = 500, Sa = 10, a2 = 50,
                   Mb = 550, b1 = 500, Sb = 10, b2 = 10,
                   SCa = 0.1, BTt2 = 10, BTt1 = 10, BTs1 = 10, BTs2 = 10))
``` 

```{r Fitted Model, fig.cap = "Fitted Model", dpi = 200}
plot(fl3.df$intensity, type = 'n', ylab = "counts", xlab = "channel") 
polygon(x = c(fl3.df$x, 257), y = c(fl3.df$intensity, 0),
                col = "lightgray", border = NA) 

## Complete model
lines(x = fl3.df$x, y = predict(m3), col = 2)

## Debris:
lines(x = fl3.df$x, y = singleCut(SCa = coef(m3)["SCa"],
                                  intensity = fl3.df$intensity), col = 3,
      lty = 2, lwd = 2)
xx <- fl3.df$x                          # awkward!!

## Internal Standard:
lines(x = fl3.df$x, y = gaussDipA(Ma = coef(m3)["Ma"], a1 = coef(m3)["a1"],
                                  Sa = coef(m3)["Sa"], a2 = coef(m3)["a2"]),
      col = 4)

## Sample:
lines(x = fl3.df$x, y = gaussDipB(Mb = coef(m3)["Mb"], b1 = coef(m3)["b1"],
                                  Sb = coef(m3)["Sb"], b2 = coef(m3)["b2"]),
      col = 5)

legend(legend = c("Internal Standard", "Sample", "Debris", "Complete Model"),
       col = c(4, 5, 3, 2), x = "topright", lty = c(1, 1, 2, 1))
```

## Extracting cell counts

Extracting cell counts requires plugging the parameter estimates back into
the formulas and integrating them.

Note that in ModFit, the results returned include values for 'All cycle
events', as well as the percent for each component, e.g. 'Diploid 41.44%,
Dip G1: 73.64%'. This means that of the 4866 total cycle events modelled,
0.4144 * 0.7364 * 4866 = 1484 are from the Diploid G1 peak. 

I can't integrate the entire model yet, as the way I have coded the debris
function doesn't work with integrate. However, we don't actually need those
values for our purposes, we can extract the components we need just fine:

```{r}
gaussAEvents <-
  integrate(function (x) coef(m3)["a1"] / (sqrt(2 * pi) * coef(m3)["Sa"]) *
                         exp(-((x - coef(m3)["Ma"])^2)/(2 *
                                                        coef(m3)["Sa"]^2)),
            0, 256)

gaussBEvents <-
  integrate(function (x) coef(m3)["b1"] / (sqrt(2 * pi) * coef(m3)["Sb"]) *
                         exp(-((x - coef(m3)["Mb"])^2)/(2 *
                                                        coef(m3)["Sb"]^2)),
            0, 256)
```

The reported values agree very closely with the results of ModFit, which
found 1484.929 and 2849.53 events, compared to my 1452.566 and 2848.404.
The larger difference in the lower peak likely reflects differences in the
way the debris is fitted.

The CVs are just σ/μ, or `coef(m3)["Sa"]/coef(m3)["Ma"]` = `r
coef(m3)["Sa"]/coef(m3)["Ma"]`. These match the estimates from ModFit
within 0.1%.

# flowFit#

### Package Examples
```{r}
biocLite("flowFit")
biocLite("flowFitExampleData")
library(flowFit)
library("flowFitExampleData")
data(QuahAndParish)
QuahAndParish[[1]]
parent.fitting.cfse <- parentFitting(QuahAndParish[[1]], "<FITC-A>")
plot(parent.fitting.cfse)
parent.fitting.cpd <- parentFitting(QuahAndParish[[1]], "<APC-A>")
plot(parent.fitting.cpd)
parent.fitting.ctv <- parentFitting(QuahAndParish[[1]], "<Alexa Fluor 405-A>")
plot(parent.fitting.ctv)

```

### flowFit in small pieces

There is an incompatibility between flowFit and our lmd files, something is
different in the keywords? I don't understand yet, but it prevents flowFit
from finding LogDecades. An ugly fix is just this:

```{r}
parent.fitting.lmd1 <- parentFitting(lmd1, "FL3.INT.LIN",
                                     logDecades =
                                       log10(lmd1@parameters@data$range[4]),
                                     verbose = TRUE) 
plot(parent.fitting.lmd1)
```

Which provides the log decades that would have been generated had the
function worked as expected. Probably not important in the end, as we don't
want to do this actual analysis, but it is a problem to overcome as I move
deeper into the flowFit code.

Note, however, that `logDecades` isn't actually used in the analysis
itself, just in the way axes are drawn after the fact:

```{r}
p.f.lmd1.LD1 <- parentFitting(lmd1, "FL3.INT.LIN",
                                     logDecades = 1, verbose = TRUE) 
plot(p.f.lmd1.LD1)
```

So that's weird. Aborting the function due to a missing value that you
don't actually use.

In any case, their algorithm is:
1. If no estimates are provided, the starting values are set as:
   - PeakPosition = mean value of the channel (assumes only one peak is
     present, as expected for this kind of experiment)
   - PeakSize = sd of the channel
   - dataRange = channel range
   These values are stored in the @parStart slot
2.  

NB: to install ncdfFlow on Debian you must download the source and install by hand:


    R CMD INSTALL ncdfFlow --configure-args="--with-hdf5='/usr/lib/x86_64-linux-gnu/hdf5/serial'"


```{r}
biocLite("ncdfFlow")
biocLite("rhdf5")

library("ncdfFlow") ## not sure if this is needed yet?

lmd <- # FCS 2.0 format
  read.ncdfFlowSet("test-data/LMD good/188-15_+_rad_2015-04-20_647.LMD",
                   dataset = 1)

lmd3 <- # FCS 3.0 format
  read.ncdfFlowSet("test-data/LMD good/188-15_+_rad_2015-04-20_647.LMD",
                   dataset = 2)
```

Note that the Gallios `.lmd` files contain the data twice, once in 10 bit
FCS2.0 format, and once in 20 bit FCS3.0 format. 

My understanding is that only the FCS2.0 format is readable via
`read.ncdfFlowset`. Calling the function without setting the `dataset`
argument will result a warning being issued, to the effect that only one of
two available datasets was read. However, attempting to access the second
dataset with `dataset = 2` produces an error, indicating something in the
second, FCS3.0 data block is improperly formatted, or is improperly parsed
by `read.ncdfFlowset`.

I have traced the problem to the `readFCStext` function in `flowCore`.
After calling `read.ncdfFlowSet`, `readFCStext` is called from
`read.FCSheader`. At this point, it returns a named vector containing
channels with names like "FS INT LIN", "FL3 INT LIN" etc. These appear to
be the names that the function uses to parse the actual flow data.
Unfortunately, these are the headers for the FCS 2 data, and the FCS 3 data
uses different labels: `FS-H`, `FL3-A` etc. So we must use the lower-level
`read.FCS` to extract the actual data, if we want to access the FCS 3.0
format.


